CannonRV: Measuring radial velocities with a data-driven spectral model

### 2016-11-06 paper/ms.tex
Analyse a, b and c

\title{Optimize the Cannon 2}
\author[1]{Jason Cao\thanks{jc6933@nyu.edu}}
\author[2]{David W. hogg\thanks{david.hogg@nyu.edu}}

\affil[1]{Department of Physics,  New York University}
\affil[2]{NYU Physics - Center for Cosmology and Particle Physics
NYU Center for Data Science
Max-Planck-Institut fuer Astronomie }

@@

The model we adopt
 \[y_{jn} = v(l_n) \cdot \theta_{j} + e_{jn} (1)\]

@@

\(l_n^{inf}\) and \( \theta_{j}\) are available after the training step.
Infer spectrum by using:
 \[y_{jn}^{inf} = v(l_n^{inf}) \cdot \theta_{j}  (2)\]

@@

The new spectrum, which is the optimized spectrum, is represented by:
\[y_{j,n}^{opt}=a\cdot y_{j-1,n}^{inf}+b\cdot y_{j,n}^{inf}+c\cdot y_{j+1,n}^{inf}(3)\]
\bigskip

We use the least chi-squared method. So our objective function is the chi-squared between the optimized spectrum and the spectrum data. 
\bigskip

$F_{obj} (a,b,c)$= \[\sum_{n=1,j=1}^{n=N,j=J} {(y_{jn} - y_{jn}^{opt})^2 \over{\sigma_{jn}^2}} \]  (4)
\bigskip

$\sigma_{jn}^2$ is the data uncertainty for star n at pixel j. The number of stellar and pixel each star are N and J. Here N and J are the number of the stellar and the number of pixel each star.  Write the objective function in the form of matrix can make it simpler. So we introduce several matrix: Y, A, C and X, which are defined as followed:

@@

\[
Y
=
\begin{bmatrix}
    Y_1&Y_2 &\dots&Y_{N}\\

@@

where
\[
 Y_n
=
\begin{bmatrix}
    y_{1,n} \\
    y_{2,n} \\
    \vdots  \\
    y_{J,n}  \\

% A
\[
A
=
\begin{bmatrix}
    A_1&A_2 &\dots&A_{N}\\

where
\[
A_n
=
\begin{bmatrix}

@@

\[
C
=
\begin{bmatrix}
    C_1&C_2 &\dots&C_{N}\\

where
\[
C_n
=
\begin{bmatrix}

@@

\(X = 
 \begin{bmatrix}

    a\\
    b \\
    c \\

\end{bmatrix}
\)

We cut matrix Y, A and C into N smaller matrix so the calculation speed is higher. 
\bigskip

Now the objective function can be described as:
$F_{obj} (a,b,c)$= \[\sum_{n=1}^{N} (Y_n-A_n X)^TC^{-1}(Y_n-A_n X) (5)\]  
\bigskip

@@

J = $\partial F_{obj}\over \partial X$ \[=\sum_{n=1}^{n=N}2\cdot [(Y_n-A_nX)^TC^{-1}]\cdot (-A_n) (6)\]   
\bigskip

@@

\[X = [\sum_{n=1}^{n=N}A_n^TC_n^{-1}A]^{-1} \cdot [\sum_{n=1}^{n=N}A^TC_n^{-1}Y_n](7)\]
\bigskip

By using (7) we can calculate a, b and c. b should be much bigger than both a and c. And a+b+c should be equal to 1.

@@

In our experiment except calculating $\theta_{j}$, we calculate a, b and c independently for each star, which means N=1 in formula(4)(5)(6)(7).

%If the optimization works, we will optimized \(\theta_{j}\) besides a, b and c, which will come soon.
%It works!

 \subsection{\label{sec:level2}Optimized $\theta_{j}$ and inferred labels}
 After optimizing the inferred spectrum, we can modify $\theta_{j}$ and get a better result. The method is to represent $\theta_j$ by using $\theta_{j-1}$ , $\theta_j$ and $\theta_{j+1}$. If $\theta_j$ is out of range, set it as zero. So the optimized $\theta$ is:
 \[\theta_{j}^{opt}=a\cdot \theta_{j-1}+b\cdot \theta_{j}+c\cdot \theta_{j+1}(8)\]
 \bigskip

 Here a,b and c are parameters obtained from spectrum optimization. Then we infer stellar labels by using $\theta_{j}^{opt}$ and generate optimized inferred labels. The data we use is the training set, which includes 548 stellar with 8575 pixel each.
 Finally, compare the chi-squared to check whether the optimization works.

@@

The labels of the four stars are:

@@

@@

%\includegraphics[width=170mm]{figure_1.png}
 \centerline{Fig.1 The optimized spectrum and the data spectrum}
\bigskip

%\includegraphics[width=170mm]{figure_2.png}
 \centerline{Fig.2 The inferred spectrum and the data spectrum}

@@

In experiment 1, we use only four stars to obtain parameters a, b and c and the result is pretty good.  What if we optimize more stars? We randomly choose about 1\% of the APOGEE DR10 and make a histogram plot. The number of stars is 639. But now parameters a, b and c are obtained from independent stars, which means the objective function only contain the chi-squared of one star and there will be 639 objective functions. And there will be 639 sets of a, b and c. Then we can plot a histogram of a, b and c to see the distribution of their values.
\bigskip

%\includegraphics[width=170mm]{figure_3.png}
\bigskip

\centerline{Fig.3 Histogram for  a, b and c}

Also. We plot a, b, c against the mean fiber number, which is the average fiber number of the spectrum we optimize. In APOGEE DR 10, the fiber number is between 1 and 640. The parameters we use are from $figure_3$

%\includegraphics[width=170mm]{figure_4.png}
\bigskip

\centerline{Fig.4 a,b,c vs Mean Fiber Number}
\bigskip

@@

%\includegraphics[width=170mm]{figure_5.png}
\bigskip

\centerline{Fig.5 Delta-chi-squared vs Mean Fiber Number}
\bigskip

>>>>

\title{CannonRV: Measuring radial velocities with a data-driven spectral model}
\author[1]{Jason Cao\thanks{jc6933@nyu.edu}}
\author[2]{David W. hogg\thanks{david.hogg@nyu.edu}}
\author[3]{Melissa Ness\thanks{ness@mpia-hd.mpg.de}}

\affil[1]{Department of Physics,  New York University}
\affil[2]{NYU Physics - Center for Cosmology and Particle Physics
NYU Center for Data Science
Max-Planck-Institut fuer Astronomie }
\affil[3]{
Max-Planck-Institut Max-Planck-Institut fuer Astronomie 17, D-69117 Heidelberg, Germany
}

@@

The model we adopt

\begin{equation}
y_{jn} = v(l_n) \cdot \theta_{j} + e_{jn}
\end{equation}

@@

\(l_n^{inf}\) and \( \theta_{j}\) are available after the training step.
Infer spectrum by using:

\begin{equation}
y_{jn}^{inf} = v(l_n^{inf}) \cdot \theta_{j} 
\end{equation}

@@

The new spectrum, which is the optimized spectrum, is represented by:

\begin{equation}
y_{j,n}^{opt}=a\cdot y_{j-1,n}^{inf}+b\cdot y_{j,n}^{inf}+c\cdot y_{j+1,n}^{inf}
\end{equation}
\bigskip

We use the least chi-squared method. So our objective function is the chi-squared between the optimized spectrum and the spectrum data. The objective function for star n is:
\bigskip

\begin{equation}
F^{obj}_n (a,b,c)= \sum_{j=1}^{j=J} {(y_{jn} - y_{jn}^{opt})^2 \over{\sigma_{jn}^2}}
\end{equation}

$\sigma_{jn}^2$ is the data uncertainty for star n at pixel j. The number of stellar and pixel each star are N and J. Here N and J are the number of the stellar and the number of pixel each star.  Write the objective function in the form of matrix can make it simpler. There will be N objective functions and N sets of a, b and c. So we introduce several matrix: $Y_n$, $A_n$, $C_n$ and $X_n$, which are defined as followed:

@@

\[
 Y_n
=
\begin{bmatrix}
    y_{1,n} \\
    y_{2,n} \\
    \vdots  \\
    y_{J,n}  \\

@@

% A

\[
A_n
=
\begin{bmatrix}

@@

\[
C_n
=
\begin{bmatrix}

@@

\begin{center}
\(X_n = 
 \begin{bmatrix}

    a_n\\
    b_n \\
    c_n \\

\end{bmatrix}
\)
\end{center}

Now the objective function can be described as:

\begin{equation}
F^{obj}_n (a_n,b_n,c_n)= (Y_n-A_n X_n)^TC^{-1}(Y_n-A_n X_n)
\end{equation}

@@

\begin{equation}
J = {\partial F^{obj}_n\over \partial X_n } = 2\cdot [(Y_n-A_nX_n)^TC^{-1}]\cdot (-A_n)
\end{equation}

@@

\begin{equation}
X_n = [\sum_{n=1}^{n=N}A_n^TC_n^{-1}A]^{-1} \cdot [\sum_{n=1}^{n=N}A^TC_n^{-1}Y_n]
\end{equation}

By using (7) we can calculate $a_n$, $b_n$ and $c_n$, which are parameters a, b and c for star n. 
b should be much bigger than both a and c. And a+b+c should be equal to 1.

@@

%It works!

\subsection{\label{sec:level2}Spectrum calibration}
\bigskip
After the optimization, we obtain many sets of a, b and c. In our theory b should be close to 1 and a, c is much smaller than b. But the fact is that sometimes b is close to -1. This is because of the spectral broadening in the measurement. The equipment is not perfect, so the resolution of the spectrum will become lower due to scattering, dispersion or other reasons. Then the spectrum will broaden and the real spectrum at pixel j is not what we obtain, it should be the spectrum at nearby pixels, which include pixel j-1 and j+1. Thus b will be close to -1 and at the same time a and c should be close to 1, which we will check in experiment 3.
\bigskip

Sometimes a or c will be close to 1. This is also because of the measurement. Even we ignore the spectral broadening, the measured spectrum may move one pixel left or one pixel right from the real spectrum. If the measured spectrum move one pixel left, then a should be close to 1 and it's similar for c. The measured spectrum moves left or right is because of the error in the measured velocity, so there will be error in measuring the Doppler effect, which leads to a spectrum moving. We will show this in experiment 4.
\bigskip

\subsection{\label{sec:level2}Optimize Teff , log g and [Fe/H] simultaneously with a, b, c}

Working on it...

@@

The labels of the four stars are:
\begin{center}

@@

 \end{center}

@@

\includegraphics[width=170mm]{figure_1.png}
 \centerline{Fig.1 The optimized spectrum and the data spectrum}
\bigskip

\includegraphics[width=170mm]{figure_2.png}
 \centerline{Fig.2 The inferred spectrum and the data spectrum}

@@

In experiment 1, we use only four stars to obtain parameters a, b and c and the result is pretty good.  What if we optimize more stars? We randomly choose about 1\% of the APOGEE DR10 and make a histogram plot. The number of stars is 907. Now parameters a, b and c are obtained from independent stars, which means the objective function only contain the chi-squared of one star and there will be 907 objective functions. And there will be 907 sets of a, b and c.

We plot a, b, c against the mean fiber number, which is the average fiber number of the spectrum we optimize. In APOGEE DR 10, the fiber number is between 1 and 640. The parameters we use are from figure 3

\includegraphics[width=170mm]{figure_3.png}
\bigskip

\centerline{Fig.3 a,b,c vs Mean Fiber Number}
\bigskip

@@

\includegraphics[width=170mm]{figure_4.png}
\bigskip

\centerline{Fig.4 Delta-chi-squared vs Mean Fiber Number}
\bigskip

\subsection{\label{sec:level2}Experiment 3}

\bigskip
Now choose some examples from the 907 stars and make some plots.

First, plot the four stars with biggest delta-chi-squared and smallest delta-chi-squared:

\includegraphics[width=170mm]{figure_5.png}
\bigskip

\centerline{Fig.5 Spectrum of the four stars with biggest delta-chi-squared}
\bigskip

\includegraphics[width=170mm]{figure_6.png}
\bigskip

\centerline{Fig.6 Spectrum of the four stars with smallest delta-chi-squared}
\bigskip

From figure 6 and 7, we find that the four stars with smallest delta-chi-squared have b closer to 1, but the four stars with biggest delta-chi-squared don't.

Also, we want to look into the spectral broadening.
Plot four stars with b closest to 1:
\bigskip

\includegraphics[width=170mm]{figure_7.png}
\bigskip

\centerline{Fig.7 Spectrum of the four stars with b closest to 1}
\bigskip

From the histogram of a, b and c, we also find that sometimes b becomes -1, which means the spectrum is broadened due to the measurement. The resolution of the spectrum becomes lower and this is what we want to calibrate.
\bigskip
Here is the plot of the four stars with b closest to -1:
\bigskip

\includegraphics[width=170mm]{figure_8.png}
\bigskip

\centerline{Fig.8 Spectrum of the four stars with b closest to -1}
\bigskip

We found a and c are close to 1, which is in accordance with our theory. 
Also, the optimized spectrum broaden when b =-1, which means the measured spectrum broaden due to the error in the measurement. This is what we need to calibrate.
\bigskip

To look deeper into the results, we also choose the four stars with biggest $a-c$ and biggest $c-a$. Then make another two plots:
\bigskip

\includegraphics[width=170mm]{figure_9.png}
\bigskip

\centerline{Fig.9 Spectrum of the four stars with biggest $a-c$}
\bigskip

\includegraphics[width=170mm]{figure_10.png}
\bigskip

\centerline{Fig.10 Spectrum of the four stars with biggest $c-a$}
\bigskip

For spectrum with $a>c$, the optimized spectrum moves right a little bit, which means the real spectrum should move left a little. It's similar for $c>a$

\subsection{\label{sec:level2}Experiment 4}

\bigskip
We also plot $Teff$ against log g. The color bar shows the mean inverse variance of these stars.

\includegraphics[width=170mm]{figure_11.png}
\bigskip

\centerline{Fig.11 $Teff$ vs $log g$}
\bigskip
